---
title: "HGNChelper Usecases: MSigDB and GEO Platforms"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

# Gene symbols in MSigDB
## Load MSigDB and Excel-modified gene symbols
Excel-mogrified gene symbols (`mog_map.csv`) was catalogued by importing all gene
symbols into Excel and exporting them in all available date formats. Also, 38,040
gene symbols used in MSigDB database is downloaded and saved in `msigdb.v7.0.symbols.gmt`.

```{r ExcellnGMT.R, warning=FALSE}
library(HGNChelper)

readGMT <- function (fname, name.column = 1) {
    if (!(name.column == 1 | name.column == 2))
        stop("name.column should be 1 or 2")
    x <- readLines(fname)
    x <- strsplit(x, split = "\t")
    output <- lapply(x, function(y) y[-1:-2])
    names(output) <- make.names(sapply(x, function(y) y[name.column]), unique = TRUE)
    return(output)
}

# gmt.files <- dir(pattern = "\\.gmt$")
# if(length(gmt.files) == 0) stop(paste("No .gmt files found in", getwd()))

# Load all .gmt files
gmt.files <- "msigdb.v7.0.symbols.gmt"
all.sigs <- lapply(gmt.files, readGMT)
all.sigs <- lapply(all.sigs, function(x) unique(unlist(x)))
# all.sigs <- lapply(all.sigs, function(x) x[!grepl("^LOC", x)]) 
names(all.sigs) <- gmt.files
all.sigs$OVERALL <- unique(do.call(c, all.sigs))

# Load Excel-mogrified gene symbols
mog <- read.csv(system.file("extdata", "mog_map.csv", package = "HGNChelper"), 
                as.is=TRUE)
```


## Update using `HGNChelper`

Fetch the up-to-date human gene symbol map using `HGNChelper::getCurrentHumanMap`.
```{r}
CurrentHumanMap = getCurrentHumanMap()
```

Update gene symbols using HGNChelper and check any Excel-mogrified gene symbol.
```{r warning=FALSE}
simplecheck <- sapply(all.sigs, function(x){
        tmp <- checkGeneSymbols(x, map = CurrentHumanMap)
        output <- c(sum(tmp$Approved)/nrow(tmp), 
                    sum(!is.na(tmp[, 3]))/nrow(tmp), 
                    any(toupper(x) %in% toupper(mog[, 2])))
        names(output) <- c("valid.frac", "valid.after.hgnchelper.frac", "any.excel")
        return(output)
    })
simplecheck <- t(simplecheck)
simplecheck

# checkGeneSymbols
tmp = checkGeneSymbols(all.sigs[[1]], map = CurrentHumanMap)

# the number of invalid gene symbols before
sum(tmp$Approved != "TRUE")   

# the number of invalid gene symbols after
sum(is.na(tmp$Suggested.Symbol))   

# Example of gene symbols that can't be corrected by HGNGhelper
unfixables <- sort(tmp[is.na(tmp[, 3]), 1])  
length(unfixables)
set.seed(1)
sample(unfixables, 10)
```

Most of unfixable cases are lncRNA. Some are withdrawn gene symbols, LOC- genes. 
(*From NCBI: "Symbols beginning with LOC. When a published symbol is not available, and orthologs have not yet been determined, Gene will provide a symbol that is constructed as 'LOC' + the GeneID. This is not retained when a replacement symbol has been identified, although queries by the LOC term are still supported. In other words, a record with the symbol LOC12345 is equivalent to GeneID = 12345. So if the symbol changes, the record can still be retrieved on the web using LOC12345 as a query, or from any file using GeneID = 12345."*)

# Gene symbols in GPL

- There are 20,716 unique platforms.    
- *platforms* directory : there are 20,716 files saved in this folder          
- *hgnc.vecs* directory : there are 2,044 files saved in this folder       

## Fetch GPL files and update using `HGNChelper`
GEO platform information is downloaded from https://www.ncbi.nlm.nih.gov/geo/browse/?view=platforms
in five `.csv` files and combined/saved as `platforms.csv`. As of March 27th, 2020, 
there are 20,716 platforms exist in GEO.

```{r eval=FALSE, echo=FALSE}
dir <- "inst/extdata/GPL_list"
gpl_files <- list.files(dir)
first <- read.csv(file.path(dir, gpl_files[1]), as.is=TRUE)
platform <- as.data.frame(matrix(ncol = ncol(first), nrow = 0))
colnames(platform) <- colnames(first)

for (i in seq_along(gpl_files)) {
  gpl = read.csv(file.path(dir, gpl_files[i]), as.is=TRUE)
  platform = rbind(platform, gpl)
}

write.csv(platform, "inst/extdata/platform.csv", row.names = FALSE)
```

Fetch all the gene symbols found in GPL,  store in column 1 of `platform.csv`.

```{r test_GPLs_1, eval=FALSE}
library(GEOquery)
all.platforms <- read.csv("platform.csv", as.is=TRUE)

# human Entrez Gene identifier symbols
library(org.Hs.eg.db)
hgnc.reference <- as.character(org.Hs.egSYMBOL)   
names(hgnc.reference) <- NULL
```

## Create `gpls_already_tested.csv` file
(This step takes long --> separate R script available as `test_GPLs.R`)
```{r test_GPLs_2, eval=FALSE}
library(HGNChelper)

raw.platform.dir <- "platforms"
dir.create(raw.platform.dir)
hgnc.vec.dir <- "hgnc.vecs"
dir.create(hgnc.vec.dir)

if (file.exists("gpls_already_tested.csv")) {
    gpls.already.tested <- read.csv("gpls_already_tested.csv", header=TRUE)
} else {
    write.table(t(c("platform", "colname", "frac.hgnc", "nrow", 
                    "valid.frac", "valid.after.hgnchelper.frac", 
                    "distribution", "submission_date")),
                file="gpls_already_tested.csv", 
                row.names=FALSE, col.names=FALSE, sep=",")
}

for (i in 1:nrow(all.platforms)){
    gpl <- all.platforms[i, 1]
    hgnc.vec.file <- paste(hgnc.vec.dir, "/", gpl, "_hgnc.vec.RData", sep="")
    
    if(exists("gpls.already.tested") && gpl %in% gpls.already.tested$platform) next
    gpldat <- try(getGEO(gpl, destdir=raw.platform.dir))   
    if(class(gpldat) == "try-error") next
    gpltable <- try(Table(gpldat))
    if(class(gpltable) == "try-error") next
    hgnc.frac <- apply(gpltable, 2, function(x) sum(unique(x) %in% hgnc.reference) / length(unique(x)))
    
    # any column containing gene symbol will have >0 value
    if(any(hgnc.frac > 0.5)) {   # updated from `hgnc.frac > 0`
      
        # assumed that there is only one 'symbol' column
        hgnc.vec <- unique(as.character(gpltable[, which.max(hgnc.frac)]))  
        # get rid of anything after a space
        hgnc.vec <- gsub("[ ].+", "", hgnc.vec)  
        # convert to ascii
        HGNChelper.output <- checkGeneSymbols(iconv(hgnc.vec, "latin1", "ASCII", ""),
                                              map = CurrentHumanMap)  
        
        # fraction of valid gene symbols BEFORE HGNChelper
        valid.frac <- sum(HGNChelper.output$Approved) / length(hgnc.vec)  
        # fraction of valid gene symbols AFTER HGNChelper
        after.HGNChelper.valid.frac <- sum(!is.na(HGNChelper.output$Suggested.Symbol)) / length(hgnc.vec)  
        
        hgnc.vec <- c(gpldat@header$distribution, gpldat@header$submission_date, hgnc.vec)
        save(hgnc.vec, file=hgnc.vec.file, compress="bzip2")
        
        info.for.file <- t(c(gpl, colnames(gpltable)[which.max(hgnc.frac)], 
                             max(hgnc.frac), nrow(gpltable), 
                             valid.frac, after.HGNChelper.valid.frac, 
                             gpldat@header$distribution, 
                             gpldat@header$submission_date))
        
        print(paste(i, gpl, length(hgnc.vec), "HGNC symbols found and saved."))
        
    } else {
       info.for.file <- t(c(gpl, colnames(gpltable)[which.max(hgnc.frac)], 
                            max(hgnc.frac), nrow(gpltable), NA, NA, 
                            gpldat@header$distribution, 
                            gpldat@header$submission_date))
       print(paste(i, gpl, ": No HGNC symbols."))
    }
    
    write.table(info.for.file, file="gpls_already_tested.csv", 
                append=TRUE, row.names=FALSE, col.names=FALSE, sep=",")
}
```


## Summary plots
### Fraction of corrected gene symbols (by original prop. of valid symbols)
```{r plot_gpls_already_tested.R}
x <- read.csv("gpls_already_tested.csv", as.is=TRUE)
x$valid.after.hgnchelper.frac <- as.numeric(x$valid.after.hgnchelper.frac)
x <- x[!is.na(x$valid.frac), ]
x <- x[x$valid.frac > 0, ]

df <- data.frame(before=cut(100*x$valid.frac, breaks=seq(0, 100, by=20)),
                 fixed=(x$valid.after.hgnchelper.frac - x$valid.frac)/(1-x$valid.frac))

par(mar=c(4,5,2,0.5))
boxplot(fixed ~ before, data=df,
        main=paste("Correcting the annotations of", 
                   nrow(df), "GEO platforms"),
        xlab="% Valid Before HGNChelper",
        ylab="Fraction of invalid \n gene symbols fixed",
        col="grey", boxwex=1.1, varwidth=TRUE)
```



### Fraction of valid gene symbols (by submission dates)

```{r valid_vs_year, fig.width=10, fig.height=4.5}
raw <- read.csv("gpls_already_tested.csv", as.is=TRUE)  # 20,713 raws
raw$valid.after.hgnchelper.frac <- as.numeric(raw$valid.after.hgnchelper.frac)
raw <- raw[!is.na(raw$valid.frac), ]   # 2,044 rows
raw <- raw[raw$valid.frac > 0.2, ]   # 2,043 rows

raw$year <- as.numeric(gsub(".+[ ]+", "", raw$submission_date))
n.years <- length(unique(raw$year))

boxplot(valid.frac ~ year, data=raw, boxwex=0.3, ylim = c(0.4, 1),
        # main="Fig 1. Valid gene symbols for submission year",
        xlab="Submission Year", ylab="Fraction of valid gene symbols", 
        names=sub("20", "'", sort(unique(raw$year))))
boxplot(valid.after.hgnchelper.frac ~ year, data=raw, ylim = c(0.4, 1),
        boxwex=0.3, xaxt='n', at=(1:n.years)+0.35, add=TRUE, col="grey")
legend("bottomleft", legend=c("Before", "After"), pch=c(0, 15), 
       col=c("black", "grey"), lty=-1, bty='n', cex=1)
```

```{r save_eps, echo=FALSE}
setEPS(width=9, height=5)
postscript("Fig1.eps")
boxplot(valid.frac ~ year, data=raw, boxwex=0.3, ylim = c(0.4, 1),
        # main="Fig 1. Valid gene symbols for submission year",
        xlab="Submission Year", ylab="Fraction of valid gene symbols", 
        names=sub("20", "'", sort(unique(raw$year))))
boxplot(valid.after.hgnchelper.frac ~ year, data=raw, ylim = c(0.4, 1),
        boxwex=0.3, xaxt='n', at=(1:n.years)+0.35, add=TRUE, col="grey")
legend("bottomleft", legend=c("Before", "After"), pch=c(0, 15), 
       col=c("black", "grey"), lty=-1, bty='n', cex=1)
dev.off()
```


The number of unique platforms (= each dot in the above plot) each year
```{r}
table(raw$year)
```


# Mouse Gene Symbol
Mouse gene symbols begin with an uppercase letter followed by all lowercase letters 
except for recessive mutations, which begin with a lowercaase letter.

```{r}
mouse = c("Pzp", "A2m", "A2mr","SEPT7", "1-Feb", "lrp1", "9/7")
checkGeneSymbols(mouse, species = "mouse")
```

HGNChelper cannot fix mouse gene alias containing uppercase letter in the middle
of gene name. For example, "E430008G22Rik" (valid symbol is "Abl1") or "AA536808"
(valid symbol is "Abl2").

```{r}
mouse = c("abl2", "AA536808", "E430008G22Rik", "AbLl", "Cpamd8", "cpamd8", "mug2", "Mug2")
checkGeneSymbols(mouse, species = "mouse")
```

```{r}
sessionInfo()
```

